---
title: "Indiana Salmonella Surveillance"
subtitle: "Summary Report"
author: "Indiana Department of Health"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    fig_caption: true
    number_sections: true
    df_print: kable
    latex_engine: xelatex
    includes:
      in_header: preamble.tex
  word_document:
    toc: true
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    df_print: kable
    reference_docx: apa7_template.docx
params:
  start_date: !r as.Date("2023-01-01")
  end_date: !r Sys.Date()
  data: NULL
  format: "pdf"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center",
  out.width = "85%",
  fig.pos = "H"  # Keep figures where they are in the text
)

# Create preamble.tex file for PDF APA style
if (params$format == "pdf") {
  cat("\\usepackage{float}
\\usepackage{setspace}
\\doublespacing
\\usepackage{geometry}
\\geometry{letterpaper, margin=1in}
\\usepackage{fancyhdr}
\\pagestyle{fancy}
\\fancyhf{}
\\fancyhead[R]{Indiana Salmonella Surveillance}
\\fancyhead[L]{Summary Report}
\\fancyfoot[C]{\\thepage}
\\renewcommand{\\headrulewidth}{0.4pt}
\\renewcommand{\\footrulewidth}{0pt}
\\usepackage{titlesec}
\\titleformat{\\section}{\\normalfont\\bfseries}{\\thesection}{1em}{}
\\titleformat{\\subsection}{\\normalfont\\bfseries}{\\thesubsection}{1em}{}
\\usepackage{caption}
\\captionsetup[figure]{font=normal,labelfont=bf,textfont=it,skip=0pt}
\\captionsetup[table]{font=normal,position=above,labelfont=bf,textfont=it,skip=0pt}
", 
  file = "preamble.tex")
}

# Load required libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(viridis)
library(lubridate)
library(RColorBrewer)

# Custom color palettes based on the dashboard's actual colors
serotype_colors <- c(
  "Agbeni" = "#BA6000",       # BROWN
  "Enteritidis" = "#1E90FF",  # Dodger blue (shown in image)
  "I 4:i:-" = "#4DAF4A",      # Green (from image)
  "Newport" = "#4baaa2",      # Dark cyan (from image)
  "Typhimurium" = "#F781BF",  # Pink (from image)
  "Braenderup" = "#4B0082",   # Indigo
  "Infantis" = "#AF1493",     # Deep pink
  "Javiana" = "#006400",      # Dark green
  "Paratyphi B var. L(+) tartrate+" = "#FF8C00", # Dark orange
  "Senftenberg" = "#9932CC",  # Dark orchid
  "Stanley" = "#FFAA99",      # PEACH
  "Saintpaul" = "#00CED1",    # Dark turquoise
  "Thompson" = "#2E8B57",     # Sea green
  "Typhi" = "#EA1",           # Crimson
  "Paratyphi A" = "#FF4500",  # Orange red
  "Paratyphi B" = "#CA6347",  # Tomato
  "Paratyphi C" = "#F07F50",  # Coral
  "Other" = "#A9A9A9"         # Dark gray
)

# Dashboard-matched colors for other elements
specimen_colors <- c(
  "Stool" = "#F17E80",    # Peach/Orange
  "Urine" = "#92b5F5",    # Bright Blue
  "Blood" = "#FF5252",    # Bright Pink/Red
  "Wound" = "#66BB6A",    # Green
  "Other" = "#11199e9e",  # Gray
  "Beef" = "#FF7043",     # Deep Orange
  "Melon" = "#26A69A",    # Teal/Green
  "Abscess" = "#EC407A",  # Pink
  "Sputum" = "#5E57C2",   # Purple
  "Gall Bladder" = "#1C6BC0" # Indigo
)

# Access the data
data <- params$data
```

\newpage

# Overview

This summary report presents Salmonella surveillance findings for Indiana from **`r format(params$start_date, "%B %d, %Y")`** to **`r format(params$end_date, "%B %d, %Y")`**. It provides visualizations and data tables from the Indiana Salmonella Surveillance Dashboard.

```{r summary-stats}
total_cases <- nrow(data)

# Calculate serotype percentages
serotype_counts <- data %>%
  count(Serotype_wgs) %>%
  arrange(desc(n)) %>%
  mutate(percentage = n / total_cases * 100)

# Top serotype
top_serotype <- serotype_counts$Serotype_wgs[1]
top_serotype_count <- serotype_counts$n[1]
top_serotype_pct <- round(serotype_counts$percentage[1], 1)

# Blood isolates
blood_count <- sum(data$Specimen == "Blood")
blood_pct <- round(blood_count / total_cases * 100, 1)

# Calculate median age
ages <- data %>%
  filter(Patient_Age_Years != "Unknown") %>%
  pull(Patient_Age_Years) %>%
  as.numeric()

median_age <- if(length(ages) > 0) round(median(ages), 1) else "Unknown"

# Sex distribution
female_count <- sum(data$Sex == "FEMALE")
female_pct <- round(female_count / total_cases * 100, 1)

male_count <- sum(data$Sex == "MALE")
male_pct <- round(male_count / total_cases * 100, 1)

# Outbreak cases
outbreak_count <- sum(data$Outbreak != "Unknown" & data$Outbreak != "N/A")
outbreak_pct <- round(outbreak_count / total_cases * 100, 1)

# Top counties
county_data <- data %>%
  filter(State == "Indiana", County != "Unknown") %>%
  count(County) %>%
  arrange(desc(n))

top_county <- county_data$County[1]
top_county_count <- county_data$n[1]
top_county_pct <- round(top_county_count / total_cases * 100, 1)
```

## Key Statistics

* **Total Cases:** `r total_cases` laboratory-confirmed Salmonella infections
* **Predominant Serotype:** `r top_serotype` (`r top_serotype_pct`% of cases)
* **Invasive Disease:** `r blood_count` blood isolates (`r blood_pct`% of cases)
* **Demographics:** Median age `r median_age` years; `r female_pct`% female, `r male_pct`% male
* **Geographic Concentration:** Highest in `r top_county` County (`r top_county_count` cases, `r top_county_pct`% of total)
* **Outbreak Association:** `r outbreak_count` cases (`r outbreak_pct`%) linked to identified outbreaks

\newpage

# Serotype Distribution

Figure 1 shows the top 10 Salmonella serotypes identified during the surveillance period.

```{r serotype-plot, fig.cap="Top 10 Salmonella Serotypes Identified in Indiana"}
# Top 10 serotypes plot
serotype_data <- data %>%
  count(Serotype_wgs) %>%
  arrange(desc(n)) %>%
  top_n(10)

ggplot(serotype_data, aes(x = reorder(Serotype_wgs, n), y = n, fill = Serotype_wgs)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = serotype_colors) +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Serotype",
    y = "Number of Cases"
  ) +
  theme(legend.position = "none")
```

\newpage

# Serotype Trends Over Time

Figure 2 illustrates monthly trends for the top 5 Salmonella serotypes during the surveillance period.

```{r serotype-trend, fig.cap="Monthly Trends of Top 5 Salmonella Serotypes"}
# Get top 5 serotypes
top_serotypes <- data %>%
  count(Serotype_wgs) %>%
  arrange(desc(n)) %>%
  head(5) %>%
  pull(Serotype_wgs)

# Prepare monthly trend data
trend_data <- data %>%
  filter(Serotype_wgs %in% top_serotypes) %>%
  mutate(Month = floor_date(Collection_Date, "month")) %>%
  count(Month, Serotype_wgs)

# Create the plot with dashboard's actual colors
ggplot(trend_data, aes(x = Month, y = n, color = Serotype_wgs, group = Serotype_wgs)) +
  geom_line(size = 0.7) +
  geom_point(size = 1.5) +
  scale_color_manual(values = serotype_colors) +
  theme_minimal() +
  labs(
    x = "Month",
    y = "Number of Cases",
    color = "Serotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

\newpage

# Demographic Patterns

## Age and Sex Distribution

Figure 3 displays the age and sex distribution of Salmonella cases.

```{r demographics, fig.cap="Age and Sex Distribution of Salmonella Cases"}
# Age and sex distribution
age_sex_data <- data %>%
  filter(Sex %in% c("MALE", "FEMALE"), Age_Group != "Unknown") %>%
  count(Age_Group, Sex)

# Order age groups correctly
age_order <- c("< 1 year", "1-5 years", "6-18 years", "19-65 years", "> 65 years")
age_sex_data$Age_Group <- factor(age_sex_data$Age_Group, levels = age_order)

ggplot(age_sex_data, aes(x = Age_Group, y = n, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("FEMALE" = "#C179A7", "MALE" = "#0072B2")) +
  theme_minimal() +
  labs(
    x = "Age Group",
    y = "Number of Cases",
    fill = "Sex"
  )
```

\newpage

## Specimen Distribution

Figure 4 shows the distribution of specimen types among Salmonella isolates.

```{r specimen-distribution, fig.cap="Specimen Type Distribution of Salmonella Isolates"}
# Specimen distribution
specimen_data <- data %>%
  count(Specimen) %>%
  arrange(desc(n)) %>%
  filter(Specimen != "Unknown")

ggplot(specimen_data, aes(x = reorder(Specimen, n), y = n, fill = Specimen)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = specimen_colors) +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Specimen Type",
    y = "Number of Cases"
  ) +
  theme(legend.position = "none")
```

\newpage

# MLST Analysis

Figure 5 presents the top 10 MLST sequence types identified, with their associated serotypes.

```{r mlst-distribution, fig.cap="Top 10 MLST Sequence Types with Associated Serotypes"}
# MLST distribution with serotype labels
mlst_data <- data %>%
  filter(MLST_ST != "Unknown") %>%
  group_by(MLST_ST) %>%
  mutate(CommonSerotype = names(which.max(table(Serotype_wgs)))) %>%
  ungroup() %>%
  count(MLST_ST, CommonSerotype) %>%
  arrange(desc(n)) %>%
  top_n(10)

# Create combined labels
mlst_data <- mlst_data %>%
  mutate(MLST_Label = paste0(MLST_ST, " (", CommonSerotype, ")"))

# Create the plot with combined labels
ggplot(mlst_data, aes(x = reorder(MLST_Label, n), y = n, fill = MLST_Label)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d(option = "A") +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "MLST Sequence Type (Serotype)",
    y = "Number of Isolates"
  ) +
  theme(legend.position = "none")
```

\newpage

# Geographic Distribution

Figure 6 shows the top 15 Indiana counties by Salmonella case count.

```{r county-distribution, fig.cap="Top 15 Indiana Counties by Salmonella Case Count"}
# County distribution
county_data_top <- county_data %>%
  top_n(15)

ggplot(county_data_top, aes(x = reorder(County, n), y = n, fill = n)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "D", direction = -1) +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "County",
    y = "Number of Cases"
  ) +
  theme(legend.position = "none")
```

\newpage

# Outbreaks

```{r outbreak-count}
# Get outbreak data
outbreak_data <- data %>%
  filter(Outbreak != "Unknown" & Outbreak != "N/A") %>%
  group_by(Outbreak) %>%
  summarize(
    Count = n(),
    `Primary Serotype` = names(which.max(table(Serotype_wgs))),
    `First Detected` = min(Collection_Date),
    `Last Detected` = max(Collection_Date),
    Duration = as.numeric(difftime(`Last Detected`, `First Detected`, units = "days"))
  ) %>%
  filter(Count >= 3) %>%
  arrange(desc(Count))

# Count significant outbreaks
sig_outbreak_count <- nrow(outbreak_data)
```

During this surveillance period, **`r sig_outbreak_count` significant outbreaks** (3+ cases) were identified.

```{r outbreak-plot, fig.cap="Significant Outbreaks (3+ Cases) by Case Count", eval=sig_outbreak_count>0}
if(sig_outbreak_count > 0) {
  # Top outbreaks
  top_outbreaks <- outbreak_data %>%
    head(10)
  
  ggplot(top_outbreaks, aes(x = reorder(Outbreak, Count), y = Count, fill = `Primary Serotype`)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = serotype_colors) +
    coord_flip() +
    theme_minimal() +
    labs(
      x = "Outbreak ID",
      y = "Number of Cases",
      fill = "Primary Serotype"
    )
}
```

\newpage

## Outbreak Timeline

Figure 8 displays the timeline of significant outbreaks during the surveillance period.

```{r outbreak-timeline, fig.cap="Timeline of Significant Outbreaks", eval=sig_outbreak_count>0}
if(sig_outbreak_count > 0) {
  # Plot outbreak timeline
  ggplot(outbreak_data, aes(x = `First Detected`, y = reorder(Outbreak, `First Detected`), 
                         color = `Primary Serotype`, size = Count)) +
    geom_point() +
    geom_segment(aes(x = `First Detected`, xend = `Last Detected`, y = Outbreak, yend = Outbreak), 
                 size = 1, alpha = 0.7) +
    scale_color_manual(values = serotype_colors) +
    scale_size_continuous(range = c(3, 8)) +
    theme_minimal() +
    labs(
      x = "Date",
      y = "Outbreak ID",
      color = "Primary Serotype",
      size = "Case Count"
    )
}
```

\newpage

# Genetic Clusters

Figure 9 visualizes genetic clusters based on Average Nucleotide Identity (ANI) analysis.

```{r ani-clusters, fig.cap="Genetic Clusters by Average Nucleotide Identity (ANI)"}
# Create ANI clusters data
ani_data <- data %>%
  filter(!is.na(ANI) & ANI != "Unknown" & !is.na(ANI_Coverage) & ANI_Coverage != "Unknown") %>%
  mutate(
    ANI = as.numeric(as.character(ANI)),
    ANI_Coverage = as.numeric(as.character(ANI_Coverage))
  ) %>%
  filter(!is.na(ANI) & !is.na(ANI_Coverage))  # Filter out any conversion errors

# Get top 10 serotypes for coloring and shapes
top_serotypes <- data %>%
  count(Serotype_wgs) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  pull(Serotype_wgs)

# Adding a category for other serotypes
ani_data <- ani_data %>%
  mutate(SerotypeCat = ifelse(Serotype_wgs %in% top_serotypes, 
    as.character(Serotype_wgs), 
    "Other"))

# Creating scatter plot of ANI vs Coverage
ggplot(ani_data, aes(x = ANI, y = ANI_Coverage, color = SerotypeCat)) +
  geom_point(alpha = 0.7, size = 3) +
  # Add vertical lines for typical boundaries
  geom_vline(xintercept = 95.0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 99.5, linetype = "dashed", color = "gray") +
  # Use dashboard colors
  scale_color_manual(values = serotype_colors, name = "Serotype") +
  labs(
    x = "Average Nucleotide Identity (ANI)",
    y = "ANI Coverage (%)",
    color = "Serotype"
  ) +
  theme_minimal() +
  # Add annotations for boundaries
  annotate("text", x = 95.0, y = max(ani_data$ANI_Coverage, na.rm = TRUE), 
    label = "Species boundary", angle = 90, vjust = 1.5, hjust = 1) +
  annotate("text", x = 99.5, y = max(ani_data$ANI_Coverage, na.rm = TRUE), 
    label = "Strain-level similarity", angle = 90, vjust = 1.5, hjust = 1)
```

\newpage

# Indiana Outbreaks Heatmap

Figure 10 presents a detailed heatmap of Salmonella serotypes by county for the top 15 counties and top 10 serotypes.

```{r serotype-county-heatmap, fig.cap="Detailed Heatmap of Salmonella Serotypes by County", fig.height=7, fig.width=8}
# Identify top serotypes and counties for the heatmap
top_serotypes_heatmap <- data %>%
  filter(Serotype_wgs != "I 4:i:-") %>%
  count(Serotype_wgs) %>%
  arrange(desc(n)) %>%
  top_n(10) %>%
  pull(Serotype_wgs)

# Get top counties
top_counties_heatmap <- data %>%
  filter(State == "Indiana", County != "Unknown") %>%
  count(County) %>%
  arrange(desc(n)) %>%
  top_n(15) %>%
  pull(County)

# Create heatmap data
heatmap_data <- data %>%
  filter(Serotype_wgs %in% top_serotypes_heatmap, 
    County %in% top_counties_heatmap,
    State == "Indiana") %>%
  count(County, Serotype_wgs) %>%
  complete(County, Serotype_wgs, fill = list(n = 0))

# Create the heatmap plot - using dashboard colors
ggplot(heatmap_data, aes(x = Serotype_wgs, y = County, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#1a1a4c", high = "#F8CF59", name = "Count") +
  theme_minimal() +
  labs(
    x = "Serotype",
    y = "County"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

---

*This summary report was generated from the Indiana Salmonella Surveillance Dashboard on `r format(Sys.Date(), "%B %d, %Y")`. It includes visualizations of surveillance data from `r format(params$start_date, "%B %d, %Y")` to `r format(params$end_date, "%B %d, %Y")`.*